Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-15 Manuscript/manuscript.Rmd
manuscript edits

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where *p* the set of annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation, which resulted in much more gaussian-shaped distance densities (Supplementary Figures 1-2 E-H).

@@

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers built on functional genomic annotations (Figure 3). TAD boundaries were called using the arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines *L* = {*GM12878*, *K562*}, at four different resolutions *R* = {*10 kb*, *25 kb*, *50 kb*, *100 kb*}. For each classifier, we considered 3 separate predictor types as the feature space including *P* = {*OC*, *OP*, *Distance*}. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including *B* = {*None*, *RUS*, *ROS*, *SMOTE*}. Each combination of the above parameters was repeated 21 times (number of autosomal chromosomes, excluding chromosome 9), *C* = {*1*, *2*, ..., *22*}. A system produced from the ensemble framework was defined as: *S* = {*c*, *r*, *p*, *b*}, where *l* $\in$ *L*, *r* $\in$ *R*, *p* $\in$ *P*, *b* $\in$ *B*, and *c* $\in$ *C*. Letting *|X|* denote the cardinality of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2016 (= 2 $\times$ 4 $\times$ 3 $\times$ 4 $\times$ 21) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods).

@@

The total number of genomic bins, which defined the dimensions of the data for modelling, was contingent upon the resolution of the contact matrix that cell line specific TADs were called on. Higher resolutions resulted in smaller unit bin widths, which corresponded to larger numbers of genomic bins (Table 1). For GM12878, there was an average of 12,214.1, 4,892.7, 2,450.9, and 1,227.5 bins created across chromosomes, for each resolution respectively. Likewise, the average number of bins, respective to resolution, were found to be similar on the K562 cell line (Supplementary Table 3).

As expected, the number of TADs called was related to resolution. Higher resolutions were found to result in a greater number of TADs called. For GM12878, an average of 375.0, 228.4, 115.1, and 50.6 TADs were called across chromosomes, for each respective resolution (Table 1). Average TADs called across chromosomes were found to be less on the K562 cell line at all resolutions (Supplementary Table 3). Although the downward trend was still present. As a result, a greater number of TADs corresponded to a greater number of TAD boundaries. For GM12878, the average number of TAD boundaries across chromosomes was found to be 1,254.6, 714.0, 364.4, and 165.3 for each resolution respectively (Table 1). Similar boundary profiles were exhibited on the K562 cell line (Supplementary Table 3).

Furthermore, results showed that average imbalance ratios were consistently and severely low across resolutions, for both cell lines. For GM12878, the average imbalance ratios were found to be 0.12, 0.17, 0.17, and 0.15 for the 10 kb, 25 kb, 50 kb, and 100 kb resolutions respectively (Table 1). Slightly smaller resolution-specific and averaged imbalance ratios were featured on the K562 cell line, yet were still consistent and invariant to resolution. Such low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not.

>>>>

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where *p* is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation, which resulted in much more gaussian-shaped distance densities (Supplementary Figures 1-2 E-H).

@@

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers built on cell line specific functional genomic annotations (Figure 3). TAD boundaries were called using the arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines *L* = {*GM12878*, *K562*}, at four different resolutions *R* = {*10 kb*, *25 kb*, *50 kb*, *100 kb*}. For each classifier, we considered 3 separate predictor types as the feature space including *P* = {*OC*, *OP*, *Distance*}. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including *B* = {*None*, *RUS*, *ROS*, *SMOTE*}. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9), *C* = {*1*, *2*, ..., *22*}. A system produced from the ensemble framework was defined as: *S* = {*c*, *r*, *p*, *b*}, where *l* $\in$ *L*, *r* $\in$ *R*, *p* $\in$ *P*, *b* $\in$ *B*, and *c* $\in$ *C*. Letting *|X|* denote the cardinality of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2016 (= 2 $\times$ 4 $\times$ 3 $\times$ 4 $\times$ 21) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

[???probably better to report medians here]
For GM12878, the average number of genomic bins created across chromosomes was 12,214.1, 4,892.7, 2,450.9, and 1,227.5 for each resolution respectively (Table 1). Likewise, the average number chromosome specific TADs called on each resolution was found to be 375.0, 228.4, 115.1, and 50.6. These TADs resulted in an average of 1254.6, 714.0, 364.4, and 165.3 chromosome specific boundaries across each resolution. Each of these quantities was find to have a significantly large ICC, indicating that the number of genomic bins, TADs called, and corresponding unique boundaries were highly correlated between the different resolutions (Table 2). On the other hand, we found that the chromosome specific imbalance ratios remained consistenly severe across all resolution, with averages of 0.12, 0.17, 0.17, and 0.15 for each resolution respectively (ICC=0.114; p=0.109). Such low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not. The K562 cell line was found to exhibit similar profiles for the above quantities (Supplementary Tables 3-4)
